Climate indicators

### 2023-02-28 DRAFT_climate_indicators.Rmd
calculate reference and sd

@@

`tg_1957` is a stars proxy object and [most commands](https://r-spatial.github.io/stars/articles/stars8.html) will not initiate any change untill later, typically I write to file and all the lazy operations are done in squeal. Therefore, I will prepare a test data set, which is smaller and which I can import to memory. Then I can perform the operations on that data set and we can see the results.

@@

```{r, warning=F}
path <- path <- ifelse(dir == "C:", 
      "R:/",
      "/data/R/")

@@

myFiles <- list.files(path2, pattern=".nc$",full.names = T)
real_temp_summer <- NULL

@@

# Get julian days
temp <- stars::read_ncdf(paste(myFiles[1]))
start_month_num <-  6
end_month_num <- 8

@@

  tic("init")
  temp <- stars::read_ncdf(paste(myFiles[i]), var="tg", proxy=F,
                           ncsub = cbind(start = c(1, 1, julian_start), 
                              count = c(NA, NA, step)))
  year_temp <- year(st_get_dimension_values(temp, "time")[1])
  print(year_temp)
  #start <- ym(paste(year_temp, start_month, sep="-"))
  #end <-  ym(paste(year_temp, end_month, sep="-"))
  #myInterval <- interval(start, end)
  #myInterval <- interval(start, start+1)
  lookup <- setNames("mean", paste0("v_", year_temp))
  st_crs(temp) <- 25833
  toc()

@@

Here's a test for the effect of splitting over more cores.
```{r, eval=F}
cl <- makeCluster(2L)
cl2 <- makeCluster(6L)
cl3 <- makeCluster(10L)

@@

```{r}
#write_stars(real_temp_summer, "P:/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/real_temp_summer.tif")
```

I might want to merge this

```{r}
real_temp_summer <- real_temp_summer %>%
  merge(name = "Year") %>%
  setNames("climate_variable")
```

```{r}
ggplot()+
  geom_stars(data = real_temp_summer[,,,1:2], downsample = c(10, 10, 0)) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~Year)
```

@@

For this step we need some more dummy data from different years. I made data sets y1 to y5 above.

```{r}
y1 <- x_summerMean
y2 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-1)) 
y3 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-2)) 
y4 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-3)) 
y5 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-4)) 
# We combine the data into one cube:
y <- c(y1, y2, y3, y4, y5) %>% 
  merge(name = "Year") %>%
  setNames("Attribute_A")
```

```{r}
ggplot() + 
  geom_stars(data = y) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~Year)
```

@@

```{r}
y_ref <- y %>%
  st_apply(c("x", "y"), FUN =  median_sd)
```

@@

```{r}
tmap_arrange(
tm_shape(y_ref)+
  tm_raster("reference_upper")
,
tm_shape(y_ref)+
  tm_raster("sd",
            palette = "-viridis")
)
```

>>>>

Checking that this is correct
```{r}
p = st_sfc(st_point(c(7039947.63, 270783.69)), crs=25833)
```

@@

`tg_1957` is a stars proxy object and [most commands](https://r-spatial.github.io/stars/articles/stars8.html) will not initiate any change until later, typically I write to file and all the lazy operations are done in squeal. Therefore, I will prepare a test data set, which is smaller and which I can import to memory. Then I can perform the operations on that data set and we can see the results.

@@

```{r, warning=F, eval=F}
path <- path <- ifelse(dir == "C:", 
      "R:/",
      "/data/R/")

@@

myFiles <- list.files(path2, pattern=".nc$",full.names = T)
# The last file, the last year, is incomplete and dont include all juliani dates that we select below
myFiles <- myFiles[-length(myFiles)]
real_temp_summer <- NULL

@@

# Get julian days
temp <- stars::read_ncdf(paste(myFiles[1]), var="tg")
start_month_num <-  6
end_month_num <- 8

@@

  tic("init")
  temp <- stars::read_ncdf(paste(myFiles[i]), var="tg", proxy=F,
                           ncsub = cbind(start = c(1, 1, julian_start), 
                              count = c(NA, NA, step)))
  year_temp <- year(st_get_dimension_values(temp, "time")[1])
  print(year_temp)
  #start <- ym(paste(year_temp, start_month, sep="-"))
  #end <-  ym(paste(year_temp, end_month, sep="-"))
  #myInterval <- interval(start, end)
  #myInterval <- interval(start, start+1)
  lookup <- setNames("mean", paste0("v_", year_temp)) # Perhaps leave out the v_ to get a numeric vector instead, which is easier to subset
  st_crs(temp) <- 25833
  toc()

@@

This takes about 20 sec per file/year, or 22 min on total. Here's a test for the effect of splitting over more cores.
```{r, eval=F}
cl <- makeCluster(2L)
cl2 <- makeCluster(6L)
cl3 <- makeCluster(10L)

@@

```{r}
names(real_temp_summer)
```

```{r}
dim(real_temp_summer)
```

```{r}
st_get_dimension_values(real_temp_summer, "Year")
```

```{r, eval=F}
saveRDS(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.rds")
write_stars(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.tiff")
```

These files are about 500 MB, but the tiff file reads in a lot quicker then the rds file, so from now on I'll just export as tiff.

```{r}
temp_median_summer <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.tiff",
                                 proxy=F)
# This is too slow:
# readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/temp_median_summer.rds")
```

Note that GTiff automatically renames the third dimension _band_ and also renames the attribute.

```{r}
st_get_dimension_values(temp_median_summer, "band")
```

I can rename them. 
```{r}
temp_median_summer <- temp_median_summer %>% 
  st_set_dimensions(names = c("x", "y", "v_YEAR")) %>%
  setNames("temperature")
dim(temp_median_summer)
```

```{r, fig.cap="Showing three random slices of the year dimension."}
ggplot()+
  geom_stars(data = temp_median_summer[,,,c(1,11,22)], downsample = c(10, 10, 0)) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~v_YEAR)
```

@@

We need to first to define a reference period and then to subset our data `temp_median_summer`. We can practice on the dummy data `y`.

```{r}
dir <- substr(getwd(), 1,2)
path <- ifelse(dir == "C:",
       "P:/", 
       "/data/P-Prosjekter2/")
path2 <- paste0(path, 
                "41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data")
myFiles <- list.files(path2, pattern=".rds", full.names = T)
y <- NULL

# Looping though the files in the directory
for(i in 1:length(myFiles)){
  temp <- readRDS(myFiles[i])
  y <- c(temp, y)
}
```

```{r}
st_get_dimension_values(y, "time")
```
```{r}
y_filtered <- y %>%
  select(Attribute_A) %>%
  filter(time %within% interval("1961-01-01", "1990-12-31"))
```

@@

```{r}
y_ref <- y_filtered %>%
  st_apply(c("x", "y"), FUN =  median_sd)
```

@@

```{r, fig.cap="Showing the upper reference levels and the standard deviation from the dummy data set."}
tmap_arrange(
tm_shape(y_ref)+
  tm_raster("reference_upper")
,
tm_shape(y_ref)+
  tm_raster("sd",
            palette = "-viridis")
)
```

Let's transfer this over to the real data. First we need to rename our dimension values and turn them back into dates.

```{r}
new_dims <- as.Date(paste0(substr(st_get_dimension_values(temp_median_summer, "v_YEAR"), 3, 6), "-01-01"))
temp_median_summer_ref <- temp_median_summer %>%
  st_set_dimensions("v_YEAR", values = new_dims)
```

Then I can filter to leave only the reference period.

```{r}
temp_median_summer_ref <- temp_median_summer_ref %>%
  filter(v_YEAR %within% interval("1961-01-01", "1990-12-31"))
st_get_dimension_values(temp_median_summer_ref, "v_YEAR")
```

And then we calculate the median and sd like above

```{r, eval=FALSE}
system.time({
cl <- makeCluster(10L)
temp_median_summer_ref <- temp_median_summer_ref %>%
  st_apply(c("x", "y"), 
           FUN =  median_sd,
           CLUSTER = cl)
stopCluster(cl)
})
```

  user  system elapsed 
  9.624   6.069  20.903

```{r}
temp_median_summer_ref <- temp_median_summer_ref %>% split("median_sd")
```

```{r}
temp_median_summer_ref <- setNames(temp_median_summer_ref, c("reference_upper", "sd"))
```

```{r, fig.cap="Showing the upper reference levels and the standard deviation from actual data of median summer temperatures."}
tmap_arrange(
tm_shape(st_downsample(temp_median_summer_ref, 10))+
  tm_raster("reference_upper")
,
tm_shape(st_downsample(temp_median_summer_ref, 10))+
  tm_raster("sd",
            palette = "-viridis")
)
```
